---
title: "Comparability of translated exam forms"
output: markdown::html_format
vignette: >
  %\VignetteIndexEntry{Comparability of translated exam forms}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

When an exam is offered in a second language, the question is whether a score
means the same thing in both. Standard DIF tools struggle when the translated
group is small, differs in ability, and many items shift in the same
direction. transDIF is built for that situation.

## Data

A source-language group of 2,000 and a translated-language group of 150 who
are 0.5 logits lower on average. Items with idioms, cultural referents,
measurement units or heavy vocabulary tend to shift in translation.

```{r}
library(transDIF)
sim <- td_simulate(n_ref = 2000, n_focal = 150, n_items = 40, seed = 5)
colSums(sim$features[-1])
```

## Calibrate each group, then link and detect DIF together

```{r}
cal <- td_calibrate(sim$responses, sim$group)
dif <- td_dif(cal)
dif
```

The linking shift (the ability difference) is the center of the densest
cluster of items, not the mean of all items, so it does not assume that DIF
cancels out. Compare with linking on
all items:

```{r}
c(true = -sim$truth$focal_mean, robust = dif$link[["c"]], all_items = dif$c_mean)
```

## Does DIF change pass rates?

```{r}
td_impact(dif, cut = 24, n_draws = 100, seed = 1)
```

## How much do the conclusions depend on the linking?

Linking assumes that DIF-free items form the largest cluster. When many
items show DIF, the data may not settle which items are DIF-free.
`td_sensitivity()` shows the results under each linking assumption, how
stable the mode is, and which items' DIF verdicts depend on the choice:

```{r}
td_sensitivity(dif, cut = 24, B = 100, n_draws = 50, seed = 1)
```

A "sensitive" verdict does not say which linking is right. It says that a
conclusion depends on an assumption the data cannot check, so report the
alternatives and send the listed items to expert review.

## What should translators look at?

```{r}
td_features(dif, sim$features)
```

## A report to start from

```{r}
cat(td_report(dif, languages = c("English", "French")))
```

With very small translated-language groups (around 50 candidates), every
method has little power to detect DIF, so treat flags as candidates for
expert review rather than conclusions.
